The helicoverpa package

### 2018-08-13 notebook/helicoverpa_notes.Rmd
Explanation of toy model

     - **JUL** [18](#u-18-JUL-2018), [23](#u-23-JUL-2018), [30](#u-30-JUL-2018)

@@

>>>>

     - **AUG** [13](#u-13-AUG-2018)
     - **JUL** [18](#u-18-JUL-2018), [23](#u-23-JUL-2018), [30](#u-30-JUL-2018)

@@

> <a name="u-18-AUG-2018">Update: 18 AUG 2018</a>

**Visual stuff looks good!**

Rose has created a very nice [visual display](http://r.rosemckeon.co.uk/helicoverpa/) of the landscape, with sliders allowing for multiple crop species and pathogen strains.

**Toy model prototype**

I have managed to complete a toy model today, which should be sufficient to produce the simulated data needed to get the key ideas across. I'm going to dump the code below, which is located in the [toy_model.R](https://github.com/bradduthie/helicoverpa/blob/dev/R/toy_model.R) data file, not yet pushed on to master. We'll need to merge the interface branch with the dev branch soon, which should be easy enough. The whole thing is in R, and includes all of the basic steps from the code structure mentioned on [30 JUL 2018](#u-30-JUL-2018), including mutation, recombination, dispersal, and pathogen and crop rotation (random or sequential). The output isn't in that helpful of a form yet, but it will get there soon. To skip the many lines of code from the [toy_model.R](https://github.com/bradduthie/helicoverpa/blob/dev/R/toy_model.R) file, [go here](#skip_AUG_code).

To quickly summarise, there are 12 parameters we can play with in the toy simulation model:

1. Number of generations to simulate (`generations`)
2. X dimension (longitude), i.e., cell number, on the landscape (`xdim`)
3. Y dimension (latitude), i.e., cell number, on the landscape (`ydim`)
4. Total number of pathogen strains that can be applied (`pathogens`)
5. Total number of crops that can be planted (`crops`)
6. Initial number of pests (`pest_init`)
7. How to rotate crops on the landscape (`crop_rotate`: where `crop_rotate = "static"` means don't rotate between generations, `crop_rotate = "rotate"` means sequence through each crop one at a time over generations, and `crop_rotate = "random"` means apply a random crop in each generation -- all done independently for each landscape cell)
8. How to rotate pathogen application on the landscape (`path_rotate`: where `path_rotate = "static"` means don't rotate between generations, `path_rotate = "rotate"` means sequence through each pathogen one at a time over generations, and `path_rotate = "random"` means apply a random pathogen in each generation -- all done independently for each landscape cell)
9. Probability that a pest will move to a new cell in the generation (`pest_move_pr`)
10. Distance a pest will move up to in any direction from its natal cell (`pest_move_dist`; note, the landscape is a torus, and there is therefore no edge -- pests that move off of one side end up on the other)
11. Number of offspring each *female* pest produces (`fecundity`)
12. Carrying capacity of pests *per cell* (`cell_K`)

Default parameter values are shown below. Note that the model is an individual-based simulation with an *extremely* simple diploid genetic architecture with two loci (one for pathogens, and one for crops). If there are $N$ pathogens, then there are $N$ alleles for pathogen resistance, and pests must have at least one allele to resist (else they die). Similarly, if there are $N$ crops, then there are $N$ alleles for being able to eat crops, and pests must have at least one to eat (else they die).

```{r}
toy_simulate_resistance <- function(generations = 20,       # Generations to sim
                                    xdim = 2,               # Land dimension 1
                                    ydim = 2,               # Land dimension 2
                                    pathogens = 1,          # Pathogen strains
                                    crops = 1,              # Crop species
                                    pest_init = 2000,       # Initial pests 
                                    crop_rotate = "static", # Crops rotated
                                    path_rotate = "static", # Pathogens rotated
                                    pest_move_pr = 0.1,     # Pest movement
                                    pest_move_dist = 1,     # Pest move distance
                                    fecundity = 8,          # Offspring per fem
                                    cell_K = 2000           # K per cell
                                    ){

    if(pest_move_dist > xdim & pest_move_dist > ydim){
        pest_move_dist <- max(c(xdim, ydim)); # Avoids error
    }
    # Start initialising the landscape and pests
    LAND <- toy_initialise_land(xdim  = xdim, ydim = ydim, 
                                pathogens = pathogens, crops = crops);
    PEST <- toy_initialise_pest(LAND, N = pest_init, p_al = pathogens, 
                                c_al = crops);
    # Start the generations
    PEST_DATA   <- NULL;
    gen         <- 1;
    while(gen < generations){
        LAND <- toy_set_crops(LAND, crops, crop_rotate);                        
        LAND <- toy_set_paths(LAND, pathogens, path_rotate);                    
        PEST <- toy_move_pest(PEST, LAND, pest_move_pr, pest_move_dist);        
        PEST <- toy_feed_pest(PEST, LAND);                                      
        if(toy_check_extinction(PEST, gen) == TRUE){ # Hate these if breaks here
            break;
        }
        PEST <- toy_kill_pest(PEST, LAND);
        if(toy_check_extinction(PEST, gen) == TRUE){
            break;
        }
        PEST <- toy_reproduce_pest(PEST, LAND, pathogens, crops, fecundity, 
                                   cell_K);
        if(toy_check_extinction(PEST, gen) == TRUE){
            break;
        }
        PEST_DATA[[gen]] <- PEST;
        gen <- gen + 1;
    }
    return(PEST_DATA);
}

summarise_pest_data <- function(PEST_DATA){
    # Density estimates
    den_list <- unlist(lapply(PEST_DATA, dim));
    den_list <- den_list[c(TRUE, FALSE)];
    return(den_list);
}

toy_check_extinction <- function(PEST, gen){
    if(is.vector(PEST) == TRUE){
        print(paste("Extinction occurred in generation", gen));
        return(TRUE);
    }
    if(dim(PEST)[1] < 4){
        print(paste("Extinction occurred in generation", gen));
        return(TRUE);
    }
    if(sum(PEST[,2] == 0) < 1 | sum(PEST[,2] == 1) < 1){
        print(paste("Extinction occurred in generation", gen));
        return(TRUE);
    }
    return(FALSE);
}

# Initialise the landscape
toy_initialise_land <- function(xdim = 2, ydim = 2, pathogens = 1, crops = 1){
    LAND      <- array(data = 0, dim = c(xdim, ydim, 3));
    p_values  <- sample(x = 1:pathogens, size = xdim * ydim, replace = TRUE);
    p_layer   <- matrix(data = p_values, ncol = xdim, nrow = ydim);
    c_values  <- sample(x = 1:crops, size = xdim * ydim, replace = TRUE);
    c_layer   <- matrix(data = c_values, ncol = xdim, nrow = ydim);
    LAND[,,2] <- p_layer; # Just occurred to me that we might want a layer for
    LAND[,,3] <- c_layer; # carrying capacity of the cell in LAND later
    return(LAND);
}

# Initialise some *very* simple pests. Each allele is just going to map one to
# one for whether a crop can be attacked or a pathogen resisted. I am not even
# going to add traits affecting dispersal in the toy version -- this can be
# uniform for all individuals
toy_initialise_pest <- function(LAND, N = 10, p_al = 1, c_al = 1){
    xdim     <- dim(LAND)[1]; # This is needed to put the pests on some
    ydim     <- dim(LAND)[2]; # landscape cell
    PEST     <- matrix(data = 0, nrow = N, ncol = 8);
    PEST[,1] <- 1:N;                                           # ID
    PEST[,2] <- sample(x = c(0, 1), size = N, replace = TRUE); # Sex
    PEST[,3] <- sample(x = 1:xdim,  size = N, replace = TRUE); # x-location
    PEST[,4] <- sample(x = 1:ydim,  size = N, replace = TRUE); # y-location
    PEST[,5] <- sample(x = 1:p_al,  size = N, replace = TRUE); # p allele 1
    PEST[,6] <- sample(x = 1:p_al,  size = N, replace = TRUE); # p allele 2
    PEST[,7] <- sample(x = 1:c_al,  size = N, replace = TRUE); # c allele 1
    PEST[,8] <- sample(x = 1:c_al,  size = N, replace = TRUE); # c allele 2
    return(PEST);
}

# Just going to make this a simple function at first, randomly changing crops
toy_set_crops <- function(LAND, crops = 1, type = "static"){
    if(crops == 1 | type == "static"){
        return(LAND);
    }
    if(type == "rotate"){
        old_crop                   <- LAND[,,3];
        new_crop                   <- old_crop + 1;
        new_crop[new_crop > crops] <- 1;
        LAND[,,3]                  <- new_crop;
    }else{ # Else it just randomises -- can think about fancier ways later
        xdim      <- dim(LAND)[1];
        ydim      <- dim(LAND)[2];
        new_cval  <- sample(x = 1:crops, size = xdim * ydim, replace = TRUE);
        new_crop  <- matrix(data = new_cval, nrow = xdim, ncol = ydim);
        LAND[,,3] <- new_crop;
    }
    return(LAND);
}

# Ditto here -- just a simple function changing pathogens
toy_set_paths <- function(LAND, paths = 1, type = "static"){
    if(paths == 1 | type == "static"){
        return(LAND);
    }
    if(type == "rotate"){
        old_path                   <- LAND[,,3];
        new_path                   <- old_path + 1;
        new_path[new_path > paths] <- 1;
        LAND[,,2]                  <- new_path;
    }else{ # Else it just randomises -- can think about fancier ways later
        xdim      <- dim(LAND)[1];
        ydim      <- dim(LAND)[2];
        new_pval  <- sample(x = 1:paths, size = xdim * ydim, replace = TRUE);
        new_path  <- matrix(data = new_pval, nrow = xdim, ncol = ydim);
        LAND[,,2] <- new_path;
    }
    return(LAND);
}

# Just a function to move pests at a given value, randomly on the landscape. The
# `prob` is just probability of moving in a time step, while `dist` is just the
# maximum distance moved in cells, in any direction.
toy_move_pest <- function(PEST, LAND, prob = 0.1, dist = 1){
    xdim                     <- dim(LAND)[1];
    ydim                     <- dim(LAND)[2];
    pests                    <- dim(PEST)[1];
    to_move                  <- rbinom(n = pests, size = 1, pr = prob);
    move_x                   <- sample(x = -dist:dist, size = pests, 
                                       replace = TRUE);
    move_y                   <- sample(x = -dist:dist, size = pests, 
                                       replace = TRUE);
    PEST[to_move == 1, 3]    <- PEST[to_move == 1, 3] + move_x[to_move == 1];
    PEST[to_move == 1, 4]    <- PEST[to_move == 1, 4] + move_y[to_move == 1];
    # The ifs below make a torus landscape so that pests don't move off of it
    # In C, this will be done with a loop that looks cleaner; the below is just
    # avoiding using a loop in R, though it would probably be fine.
    if(length(PEST[PEST[,3] < 1, 3]) > 0){ 
        PEST[PEST[,3] < 1, 3] <- PEST[PEST[,3] < 1, 3] + xdim;
    }
    if(length(PEST[PEST[,3] > xdim, 3]) > 0){
        PEST[PEST[,3] > xdim, 3] <- PEST[PEST[,3] > xdim, 3] - xdim;
    }
    if(length(PEST[PEST[,4] < 1, 4]) > 0){
        PEST[PEST[,4] < 1, 4] <- PEST[PEST[,4] < 1, 4] + ydim;
    }
    if(length(PEST[PEST[,4] > ydim, 4]) > 0){
        PEST[PEST[,4] > ydim, 4] <- PEST[PEST[,4] > ydim, 4] - ydim;
    }
    return(PEST);
}

# Need to now feed the pests and remove those that don't feed (crop unavailable)
toy_feed_pest <- function(PEST, LAND){
    # I'm just going to use a loop here, else matching to LAND cells is rough
    pests <- dim(PEST)[1];
    eaten <- rep(x = 0, times = pests); 
    for(i in 1:pests){
        x_loc <- PEST[i, 3];
        y_loc <- PEST[i, 4];
        food  <- LAND[x_loc, y_loc, 3];
        if(PEST[i, 7] == food | PEST[i, 8] == food){
            eaten[i] <- 1;
        }
    }
    PEST <- PEST[eaten == 1,]; # You don't eat, you don't live
    return(PEST);
}

# Then need to have pests resist pathogens, kill those that cannot
toy_kill_pest <- function(PEST, LAND){
    # I'm just going to use a loop here, else matching to LAND cells is rough
    pests    <- dim(PEST)[1];
    survived <- rep(x = 0, times = pests);
    for(i in 1:pests){
        x_loc <- PEST[i, 3];
        y_loc <- PEST[i, 4];
        patho <- LAND[x_loc, y_loc, 2];
        if(PEST[i, 5] == patho | PEST[i, 6] == patho){
            survived[i] <- 1;
        }
    }
    PEST <- PEST[survived == 1,]; # You don't eat, you don't live
    return(PEST);
}

# Then, reproduce pests that are left
toy_reproduce_pest <- function(PEST, LAND, pa, cr, births = 2, K = 100){
    x_dim           <- dim(LAND)[1];
    y_dim           <- dim(LAND)[2];
    offspring       <- NULL;
    total_offspring <- 0; # This and the below are just to avoid an rbind(),
    cell            <- 1; # which is a massive memory sink
    lst_ID          <- max(PEST[,1]);
    for(xloc in 1:x_dim){ # Double loop is just much cleaner here
        for(yloc in 1:y_dim){
            locals     <- which(PEST[,3] == xloc & PEST[,4] == yloc);
            local_offs <- NULL;
            if( length(locals) > 1 ){
                local_offs <- toy_breed_locals(PEST, locals, births, K, lst_ID);
            }
            if(length(local_offs) > 0){
                lst_ID            <- lst_ID + dim(local_offs)[1];
                total_offspring   <- total_offspring + dim(local_offs)[1];
                offspring[[cell]] <- local_offs;
            }else{
                offspring[[cell]] <- NULL;
            }
            cell              <- cell + 1;
        }
    }
    offspring <- build_new_pest(offspring, total_offspring, pa, cr);
    return(offspring);
}

# Need to recombine the genomes correctly in the offspring
toy_breed_locals <- function(PEST, locals, births, K, last_ID){
    loc_PEST <- PEST[locals,]; # Need two of each sex (allee effect)
    if(sum(loc_PEST[,2] == 0) < 2 | sum(loc_PEST[,2] == 1) < 2){
        return(NULL);
    }
    females       <- loc_PEST[loc_PEST[,2] == 0,];
    males         <- loc_PEST[loc_PEST[,2] == 1,];
    new_offs      <- dim(females)[1] * floor(births);  
    offspring     <- matrix(data = 0, nrow = new_offs, ncol = 8);     
    offspring[,1] <- (last_ID + 1):(last_ID + new_offs);
    offspring[,2] <- sample(x = c(0, 1), size = new_offs, replace = TRUE);
    offspring[,3] <- loc_PEST[1, 3];
    offspring[,4] <- loc_PEST[1, 4];
    # Below grabs all of the alleles from females and males
    p_fem_alleles <- c(females[,5], females[,6]);
    p_mal_alleles <- c(males[,5], males[,6]);
    c_fem_alleles <- c(females[,7], females[,8]);
    c_mal_alleles <- c(males[,7], males[,8]);
    # Now add them to the offspring randomly, one from female and one from male
    for(i in 1:new_offs){
        if(runif(n = 1) < 0.5){
            offspring[i, 5] <- sample(x = p_fem_alleles, size = 1);
            offspring[i, 6] <- sample(x = p_mal_alleles, size = 1);
        }else{
            offspring[i, 5] <- sample(x = p_mal_alleles, size = 1);
            offspring[i, 6] <- sample(x = p_fem_alleles, size = 1);            
        }
        if(runif(n = 1) < 0.5){
            offspring[i, 7] <- sample(x = c_fem_alleles, size = 1);
            offspring[i, 8] <- sample(x = c_mal_alleles, size = 1);
        }else{
            offspring[i, 7] <- sample(x = c_mal_alleles, size = 1);
            offspring[i, 8] <- sample(x = c_fem_alleles, size = 1);            
        }
    }
    if(dim(offspring)[1] > K){
        offspring <- offspring[1:K,];
    }
    return(offspring);
}

# Merge the different layers into a new pest array
build_new_pest <- function(offspring, total_offspring, pa, cr, mutation = 0.01){
    new_PEST   <- matrix(data = 0, nrow = total_offspring, ncol = 8);
    start_row  <- 1; # Again, avoids an rbind memory issue
    for(cell in 1:length(offspring)){
        if(length(offspring[[cell]]) > 0){
            cell_offs       <- dim(offspring[[cell]])[1];
            rows            <- start_row:(start_row + cell_offs - 1);
            new_PEST[rows,] <- offspring[[cell]];
            start_row       <- start_row + cell_offs;
        }
    }
    mu5 <- which(rbinom(n = total_offspring, size = 1, pr = mutation) == 1);
    mu6 <- which(rbinom(n = total_offspring, size = 1, pr = mutation) == 1);
    mu7 <- which(rbinom(n = total_offspring, size = 1, pr = mutation) == 1);
    mu8 <- which(rbinom(n = total_offspring, size = 1, pr = mutation) == 1);
    new_PEST[mu5, 5] <- sample(x = 1:cr, size = length(mu5), replace = TRUE);
    new_PEST[mu6, 6] <- sample(x = 1:cr, size = length(mu6), replace = TRUE);
    new_PEST[mu7, 7] <- sample(x = 1:pa, size = length(mu7), replace = TRUE);
    new_PEST[mu8, 8] <- sample(x = 1:pa, size = length(mu8), replace = TRUE);
    return(new_PEST);
}
```

<a name="skip_AUG_code">Okay</a>, so what can the above code do? I'll get some fancy results later, but for now, let's just look at many simulations across different `pathogens` and `crops` to see how it affects pest global density. Below I run and plot the situation with just one pathogen and one crop. Note that when there is only 1 pathogen or crop, set rotation to `static`.

```{r}
sim1 <- toy_simulate_resistance(pathogens = 1, crops = 1, cell_K = 2000, 
                                pest_init = 2000, fecundity = 8, 
                                generations = 20, pest_move_pr = 0.1, 
                                crop_rotate = "static", path_rotate = "static");
dat1 <- summarise_pest_data(sim1);
plot(x = 1:length(dat1), y = dat1, type = "b", pch = 20, lwd = 2, cex.lab = 1.5,
     cex.axis = 1.5, ylim = c(0, 8000), xlim = c(1, 19), xlab = "Generation",
     ylab = "Pest density");
```

Next, we can see what happens when we instead have 3 pathogens under otherwise identical conditions. In each generation, one of three pathogens is randomly applied to each landscape cell.

```{r}
sim2 <- toy_simulate_resistance(pathogens = 3, crops = 1, cell_K = 2000, 
                                pest_init = 2000, fecundity = 8, 
                                generations = 20, pest_move_pr = 0.1, 
                                crop_rotate = "static", path_rotate = "random");
dat2 <- summarise_pest_data(sim2);
plot(x = 1:length(dat2), y = dat2, type = "b", pch = 20, lwd = 2, cex.lab = 1.5,
     cex.axis = 1.5, ylim = c(0, 8000), xlim = c(1, 19), xlab = "Generation",
     ylab = "Pest density");
```

Next, we can see what happens when we instead have 3 pathogens under otherwise identical conditions. In each generation, one of three pathogens is randomly applied to each landscape cell.

```{r}
sim3 <- toy_simulate_resistance(pathogens = 3, crops = 3, cell_K = 2000, 
                                pest_init = 2000, fecundity = 8, 
                                generations = 20, pest_move_pr = 0.1, 
                                crop_rotate = "random", path_rotate = "random");
dat3 <- summarise_pest_data(sim3);
plot(x = 1:length(dat3), y = dat3, type = "b", pch = 20, lwd = 2, cex.lab = 1.5,
     cex.axis = 1.5, ylim = c(0, 8000), xlim = c(1, 19), xlab = "Generation",
     ylab = "Pest density");
```

As is evident, pathogen and crop rotation decreases pest density. More on this later, including diversity results, but this demonstrates a working toy model. As the heterogeneity of pathogens and crops increases, pest resistance seems to go down.
